Function `cheering()` in the `hyper2` package

### 2025-02-16 inst/cheering.Rmd
cheering() in use

---
title: "A generalization of Plackett-Luce likelihoods to incorporate non-independence of competitors"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
---

@@

To cite the `hyper2` package in publications, please use @hankin2017_rmd.
Here I modify standard Plackett-Luce likelihood functions to include
an element of non-independence among the competitors.  There seem to
be a few distinct statistical models for nonindependence:

* Runner $i$ might wish to support $j$.  If runner $i$ finishes, once
  his rank is established he might be able to help $j$ in some way.
  He might cheer his still-running friend, spurring him to run faster
  and augmenting his competitive strength.

* In general, we can consider $n$ ordered pairs $(i_r,j_r)$, $1\leq
  r\leq n$, with runner $i_1$ helping runner $j_1$, runner $i_2$
  helping $j_2$, and so on.  One simple case would be equivalence
  classes of $[n]=\{1,2,\ldots,n\}$ with each equivalence class
  comprising mutually supporting runners.  In this model, with a pair
  of mutually supporting runners only the higher-placed competitor can
  affect the other.  Each equivalence class might have a different
  supportive proclivity.

* Runner $i$ might hate runner $j$.  If $i$ finishes, he might expend
  his spare energy hindering $j$ (by throwing stones or booing, for
  example).  Note that equivalence classes of mutual hatred might
  exist.

* Also, the help or hindrance might have a finite lifetime.  Under one
  model, if $i$ finishes then he can stay helping (or hindering) $j$
  wherever $j$ places.  Alternatively, suppose $i$ places $n$th.  Then
  perhaps he can affect $j$ only place $n+1$, and after a competitor
  finishes in place $n+1$, then $i$ can exert no further influence.

* The help or hindrance discussed above might be contingent on
  placing: perhaps only the winnner (that is, first-placing
  competitor) would be able to influence other competitors.  Or maybe
  competitors with any placing could influence other competitors.

* Runners $i$ and $j$ might elect to cross the finishing line holding
  hands, with the apparent tie broken randomly.  This would be a
  symmetrical relationship.  In general, we could consider
  hand-holding equivalence classes.

* We might have a resource allocation problem, with runners having a
  certain [possibly unknown] amount of "strength" but also the ability
  to expend this strength on objectives other than ranking in the
  race.  A runner might be able to expend just enough energy to remain
  slightly ahead of a rival, reserving energy for other purposes.  If
  runner $i$ wishes to beat runner $j$, and has sufficient resources
  to do so, then runners $i$ and $j$ will finish consecutively with
  $i$ ahead of $j$.

# Start simple: three runners

The simplest nontrivial case is three runners, $\left\lbrace
1,2,3\right\rbrace$ where runner 1 helps runner 2 with multiplier
$\lambda$ [and that is the only non PL interaction].  Note that this
is _not_ a symmetric interaction: 1 helps 2, but 2 does not help 1.
The model is that 1 helps 2 if and only if 1 has finished and 2 has
not.

123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$

132: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$

213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_3}$

231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{p_1+p_3}$

312: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_2}$

321: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_2}{p_1+p_2}$

(Above, the probabilities clearly sum to 1).  Next simplest: 1 and 2
mutually support each other with $\lambda$:

123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$

132: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$

213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{\lambda p_1}{p_1+p_3}$

231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_1+p_3}$

312: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_2}$

321: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_2}{p_1+p_2}$

Now we retain the mutual support between 1 and 2 but allow an
asymmetry: 1 helps 2 with $\lambda$ as before, but now in addition, 2
helps 1 with $\mu$:

123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$

132: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$

213: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{\mu p_1}{\mu p_1+p_3}$

231: $\frac{p_2}{p_1+p_2+p_3}\cdot\frac{p_3}{\mu p_1+p_3}$

312: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_2}$

321: $\frac{p_3}{p_1+p_2+p_3}\cdot\frac{p_2}{p_1+p_2}$

Now how about 1 must beat 2.  He does not care whether he comes first
or second, only that he must beat 2.  Player 1 has sufficient energy
to win if he wants, but is subject to some sort of resource constraint
(money, perhaps) and wishes to expend only enough money to beat 2.  So
conventional P-L likelihoods are modified for 1, but only if 2 is
_still running_.

123: $\frac{\lambda p_1}{\lambda p_1+p_2+p_3}\cdot\frac{p_2}{p_2+p_3}$

132: $\frac{\lambda p_1}{\lambda p_1+p_2+p_3}\cdot\frac{p_3}{p_2+p_3}$

213: $\frac{p_2}{\lambda p_1+p_2+p_3}\cdot\frac{p_1}{p_1+p_3}$

231: $\frac{p_2}{\lambda p_1+p_2+p_3}\cdot\frac{p_3}{p_1+p_3}$

312: $\frac{p_3}{\lambda p_1+p_2+p_3}\cdot\frac{\lambda p_1}{\lambda p_1+p_2}$

321: $\frac{p_3}{\lambda p_1+p_2+p_3}\cdot\frac{p_2}{\lambda p_1+p_2}$

## Four runners

Now we have four runners $\left\lbrace 1,2,3,4\right\rbrace$, again
with the only interaction being runner 1 helping runner 2.  Thus, in a
situation where $p_1$ has finished and $p_2$ is still running, we have
$p_2\longrightarrow\lambda p_2$ [although the $\lambda$ cancels if 2
is the only competitor still running, that is, if *all* the remaining
competitors are helped]

1234: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_3+p_4}\cdot\frac{p_3}{p_3+p_4}$

1243: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_3+p_4}\cdot\frac{p_4}{p_3+p_4}$

1324: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_4}$

1342: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3+p_4}\cdot\frac{p_4}{\lambda p_2+p_4}$

1324: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_4}$

1423: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{p_1+\lambda p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$

1432: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{p_1+\lambda p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3}$

2xxx: P-L [no help given by 2 to 1]

3124: $\frac{p_3}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_4}\cdot\frac{\lambda p_2}{\lambda p_2+p_4}$

3142: $\frac{p_3}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_4}\cdot\frac{p_4}{\lambda p_2+p_4}$

32xx: P-L

3412: P-L [1 helps 2 but when 1 crosses the finishing line, only 2 is
still running; there are no non-helped competitors]

4123: $\frac{p_4}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$

4132: $\frac{p_4}{p_1+p_2+p_3+p_4}\cdot\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_3}{\lambda p_2+p_3}$

42xx: P-L

43xx: P-L

Now 1 and 2 mutually support one another by $\lambda$.  Here I will
put the last term in, to show explicitly that the $\lambda$ term
remains in the expressions for numerator and denominator.

1234: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
p_2+p_3+p_4}\cdot\frac{p_3}{p_3+p_4}\cdot\frac{p_4}{p_4}$

1243: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
p_2+p_3+p_4}\cdot\frac{p_4}{p_3+p_4}\cdot\frac{p_3}{p_3}$

1324: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda
p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
p_2+p_4}\cdot\frac{p_4}{p_4}$

1342: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_3}{\lambda
p_2+p_3+p_4}\cdot\frac{p_4}{\lambda p_2+p_4}\cdot\frac{\lambda
p_2}{\lambda p_2}$

1423: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{\lambda
p_2+p_3+p_4}\cdot\frac{\lambda p_2}{\lambda
p_2+p_3}\cdot\frac{p_3}{p_3}$

1432: $\frac{p_1}{p_1+p_2+p_3+p_4}\cdot\frac{p_4}{\lambda
p_2+p_3+p_4}\cdot\frac{p_3}{\lambda p_2+p_3}\cdot\frac{\lambda
p_2}{\lambda p_2}$

2134: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
p_1+p_3+p_4}\cdot\frac{p_3}{p_3+p_4}\cdot\frac{p_4}{p_4}$

2143: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
p_1+p_3+p_4}\cdot\frac{p_4}{p_3+p_4}\cdot\frac{p_3}{p_3}$

2314: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_3}{\lambda
p_1+p_3+p_4}\cdot\frac{\lambda p_1}{\lambda
p_1+p_4}\cdot\frac{p_4}{p_4}$

2341: $\frac{p_2}{p_1+p_2+p_3+p_4}\cdot\frac{\lambda p_3}{\lambda
p_1+p_3+p_4}\cdot\frac{p_4}{\lambda p_1+p_4}\cdot\frac{\lambda
p_1}{\lambda p_1}$

We will calculate a Plackett-Luce likelihood function for four runners
$\left\lbrace 1,2,3,4\right\rbrace$ but with a reified Bradley-Terry
monster that corresponds to having subsets of mutual friends who
support one another.  For example, we might have runners 1 and 2 being
friends, and runners 3 and 4 being friends: we would represent this as
$\left\lbrace\left\lbrace 1,2\right\rbrace,\left\lbrace
3,4\right\rbrace\right\rbrace$.

The runner metaphor is not the only one.  We can apply the resulting
likelihood functions to other cases including the red bus-blue bus
problem (see `inst/red_bus_blue_bus.Rmd` for more discussion).

First do three runners, and suppose the equivalence classes are
$\left\lbrace\left\lbrace 1,2\right\rbrace,\left\lbrace
3\right\rbrace\right\rbrace$.

 123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_2+M}{p_2+p_3+M}\qquad\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$

 132: $\frac{p_1}{p_1+p_3+p_2}\cdot\frac{p_3}{p_3+p_2+M}\qquad\frac{p_1}{p_1+p_3+p_2}\cdot\frac{p_3}{p_3+\lambda p_2}$

 213: $\frac{p_2}{p_2+p_1+p_3}\cdot\frac{p_1+M}{p_1+p_3+M}\qquad\frac{p_2}{p_2+p_1+p_3}\cdot\frac{\lambda p_1}{\lambda p_1+p_3}$

 231: $\frac{p_2}{p_2+p_3+p_1}\cdot\frac{p_3}{p_3+p_1+M}\qquad\frac{p_2}{p_2+p_1+p_3}\cdot\frac{\lambda p_1}{p_3+\lambda p_1}$

 312: $\frac{p_3}{p_3+p_1+p_2}\cdot\frac{p_1}{p_2+p_1}\qquad\frac{p_3}{p_3+p_1+p_2}\cdot\frac{\lambda p_1}{p_3+p_1}$

 321: $\frac{p_3}{p_3+p_2+p_1}\cdot\frac{p_2}{p_2+p_1}\qquad\frac{p_3}{p_3+p_2+p_1}\cdot\frac{p_2}{p_2+p_1}$

Now four runners with classes $\left\lbrace\left\lbrace
1,2\right\rbrace,\left\lbrace 3,4\right\rbrace\right\rbrace$.

 1234: $\frac{1}{1+2+3+4}\cdot\frac{2+M}{2+3+4+M}\cdot\frac{3}{3+4}$

 1243: $\frac{1}{1+2+4+3}\cdot\frac{2+M}{2+4+3+M}\cdot\frac{4}{4+3}$

 1324: $\frac{1}{1+3+2+4}\cdot\frac{3}{3+2+4+M}\cdot\frac{2}{2+4+M}$

 1342: $\frac{1}{1+3+4+2}\cdot\frac{3}{3+4+2+M}\cdot\frac{4+M}{2+4+M}$

 1423: $\frac{1}{1+4+2+3}\cdot\frac{4}{4+2+3+M}\cdot\frac{2}{2+3}$

 1432: $\frac{1}{1+4+3+2}\cdot\frac{4}{4+3+2+M}\cdot\frac{3}{2+3}$

 2134: $\frac{2}{2+1+3+4}\cdot\frac{1+M}{1+3+4+M}\cdot\frac{3}{3+4}$

 2143: $\frac{2}{2+1+4+3}\cdot\frac{1+M}{1+4+3+M}\cdot\frac{4}{4+3}$

 2314: $\frac{2}{2+3+1+4}\cdot\frac{3}{3+1+4+M}\cdot\frac{1}{1+4+M}$

 2341: $\frac{2}{2+3+4+1}\cdot\frac{3}{3+4+1+M}\cdot\frac{4+M}{1+4+M}$

 2413: $\frac{2}{2+4+1+3}\cdot\frac{4}{4+1+3+M}\cdot\frac{1}{1+3}$

 2431: $\frac{2}{2+4+3+1}\cdot\frac{4}{4+3+1+M}\cdot\frac{3}{1+3}$

etc.

Now 123 tethered, 45 tethered:

12345: $\frac{1}{1+2+3+4+5}\cdot\frac{2+M}{2+3+4+5+M}\cdot\frac{3+M}{3+4+5+M}\cdot\frac{4}{4+5}$

12354: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{3+M}{3+5+4+M}\cdot\frac{5}{5+4}$

12435: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{4}{3+5+4+M}\cdot\frac{3}{3+5+M}$

12453: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{4}{3+5+4+M}\cdot\frac{5+M}{5+3+M}$

12534: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+5+3+4+M}\cdot\frac{5}{3+5+4+M}\cdot\frac{3}{3+4+M}$

12543: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+5+4+3+M}\cdot\frac{5}{3+5+4+M}\cdot\frac{4+M}{4+3+M}$

Might be better to consider ${5\choose 2\,3}$ arrangements of
$aaabb\ldots bbaaa$.  FOLLOWING IS NOT RIGHT: THE PROBABILITY MODEL IS
NOT PRECISELY DEFINED.

aaabb: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{a+M}{a+b+b+M}\cdot\frac{b}{b+b+M}$

aabab: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{b}{a+b+b+M}\cdot\frac{a}{a+b+M}$

aabba: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{b}{a+b+b+M}\cdot\frac{b+M}{b+a+M}$

abaab: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{a}{a+a+b+M}\cdot\frac{a+M}{a+b+M}$

ababa: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{a}{a+a+b+M}\cdot\frac{b}{b+a+M}$

abbaa: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{b+M}{b+a+a+M}\cdot\frac{a}{a+a+M}$

baaab: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{a+M}{a+a+b+M}\cdot\frac{a+M}{a+b+M}$

baaba: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{a+M}{a+a+b+M}\cdot\frac{b}{a+b+M}$

babaa: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{b}{b+a+a+M}\cdot\frac{a}{a+a}$

bbaaa: $\frac{b}{a+a+a+b+b}\cdot\frac{b+M}{a+a+a+b+M}\cdot\frac{a}{a+a+a}\cdot\frac{a}{a+a}$

```{r,label=definefunctionf}
f <- function (a,b,M){
return(
  factorial(3)*factorial(2)*(
   a/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(a+M)/(a+b+b+M)*(b  )/(b+b+M) +
   a/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b  )/(a+b+b+M)*(a  )/(a+b+M) +
   a/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b  )/(a+b+b+M)*(b+M)/(b+a+M) +
   a/(a+a+a+b+b)*(b  )/(a+a+b+b+M)*(a  )/(a+a+b+M)*(a+M)/(a+b+M) +
   a/(a+a+a+b+b)*(b  )/(a+a+b+b+M)*(a  )/(a+a+b+M)*(b  )/(b+a+M) +
   a/(a+a+a+b+b)*(b  )/(a+a+b+b+M)*(b+M)/(b+a+a+M)*(a  )/(a+a+M) +
   b/(a+a+a+b+b)*(a  )/(a+a+a+b+M)*(a+M)/(a+a+b+M)*(a+M)/(a+b+M) +
   b/(a+a+a+b+b)*(a  )/(a+a+a+b+M)*(a+M)/(a+a+b+M)*(b  )/(a+b+M) +
   b/(a+a+a+b+b)*(a  )/(a+a+a+b+M)*(b  )/(b+a+a  )*(a  )/(a+a  ) +
   b/(a+a+a+b+b)*(b+M)/(a+a+a+b+M)*(a  )/(a+a+a  )*(a  )/(a+a  )
   ) 
   )
}
```

```{r evalat44}
f(4.4,3.3, 0.2)*12
f(4.4,3.3, 0.0)*12
```

```{tikz, label=tikzpartprobtree1,fig.cap = "Partial probability tree for five competitors $a-e$ with $a$ supporting $b$, hyper3 approach", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw (0,0) -- (2,3); 
\draw (0,0) -- (2,2);
\draw (0,0) -- (2,1);
\draw (0,0) -- (2,0);
\draw (0,0) -- (2,-1);
\node at (0,2.5) (eq1) {$a\longrightarrow b$};

\newcommand{\lb}{\lambda b }
\newcommand{\lc}{\lambda c }
\newcommand{\lee}{\lambda e }
\node at (6, 5)  {$\frac{\lb}{\lb+c+d+e}$};
\node at (6, 4)  {$\frac{c}{\lb+c+d+e}$};
\node at (6, 3)  {$\frac{d  }{\lb+c+d+e}$};
\node at (6, 2)  {$\frac{e  }{\lb+c+d+e}$};

\fill (7.4, 5) circle[radius=2pt];  % ab finishes; paths cde
\draw (7.4,5) -- (8.4,6);
\draw (7.4,5) -- (8.4,5);
\draw (7.4,5) -- (8.4,4);

\node at (7.9,5.7) {$c$};
\node at (7.9,5.2) {$d$};
\node at (7.9,4.7) {$e$};
\node at (9.3, 6) {$\frac{c}{c+d+e}$};
\node at (9.3, 5) {$\frac{d  }{c+d+e}$};
\node at (9.3, 4) {$\frac{e  }{c+d+e}$};

\fill (10.2, 5) circle[radius=2pt];
\draw (10.2, 5) -- (11.2,5.3);
\node at (11.8, 5.3) {$\frac{c}{c+e}$};
\node at (10.7,5.3) {$c$};

\draw (10.2, 5) -- (11.2,4.7);
\node at (11.8, 4.7) {$\frac{e}{c+e}$};
\node at (10.7,4.7) {$e$};

\fill (10.2, 4) circle[radius=2pt];
\draw (10.2, 4) -- (11.2,3.9);
\node at (11.8, 3.9) {$\frac{c}{c+e}$};
\node at (10.7,4.1) {$c$};

\draw (10.2, 4) -- (11.2,3.1);
\node at (11.8, 3.1) {$\frac{e}{c+e}$};
\node at (10.7,3.7) {$e$};

\draw (4, 0) -- (5,1);
\node at (6, 1)  {$\frac{a}{a+b+c+e}$};

\draw (4, 0) -- (5,0);
\node at (6, 0)  {$\frac{b}{a+b+c+e}$};

\draw (4, 0) -- (5,-1);
\node at (6, -1)  {$\frac{c}{a+b+c+e}$};

\draw (4, 0) -- (5,-2);
\node at (6, -2)  {$\frac{e}{a+b+c+e}$};

\node at (11.8, 0) {$\frac{\lb}{\lb+c}$};
\draw (10.2, -1) -- (11.2,-1);

\node at (11.8, -1) {$\frac{c}{\lb+c}$};

\draw (10.2, -2) -- (11.2,-1.6);
\node at (11.8, -1.6) {$\frac{a}{a+c}$};

\draw (10.2, -2) -- (11.2,-2.2);
\node at (11.8, -2.2) {$\frac{c}{a+c}$};

\draw (10.2, -3) -- (11.2,-2.9);
\node at (11.8, -2.9) {$\frac{a}{a+b}$};

\draw (10.2, -3) -- (11.2,-3.4);
\node at (11.8, -3.4) {$\frac{b}{a+b}$};

\node at (9.3, 2) {$\frac{\lb}{\lb+c+e}$};
\node at (9.3, 1) {$\frac{  c}{\lb+c+e}$};
\node at (9.3, 0) {$\frac{e  }{\lb+c+e}$};

\fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
\draw (10.2, 1) -- (11.2,2);
\node at (11.8, 2) {$\frac{\lb}{\lb+e}$};
\draw (10.2, 1) -- (11.2,1);
\node at (11.8, 1) {$\frac{e}{\lb+e}$};

```{tikz, label=tikzpic2, fig.cap = "Partial probability tree for five competitors $a-e$ with mutually supporting  group $abc$, hyper3 approach.  If any competitor in the set {$a,b,c$} finishes, they lend their strength to the other competitors in the set who are still running", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw (0,0) -- (2,3); 
\draw (0,0) -- (2,2);
\draw (0,0) -- (2,1);
\draw (0,0) -- (2,0);
\draw (0,0) -- (2,-1);
\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}$};

\newcommand{\la}{\lambda a }
\newcommand{\lb}{\lambda b }
\newcommand{\lc}{\lambda c }
\newcommand{\lee}{\lambda e }
 \node at (6, 5)  {$\frac{\lb}{\lb+\lc+d+e}$};
 \node at (6, 4)  {$\frac{\lc}{\lb+\lc+d+e}$};
 \node at (6, 3)  {$\frac{d  }{\lb+\lc+d+e}$};
 \node at (6, 2)  {$\frac{e  }{\lb+\lc+d+e}$};

 \fill (7.4, 5) circle[radius=2pt];  % ab finishes; paths cde
 \draw (7.4,5) -- (8.4,6);
 \draw (7.4,5) -- (8.4,5);
 \draw (7.4,5) -- (8.4,4);

 \node at (7.9,5.7) {$c$};
 \node at (7.9,5.2) {$d$};
 \node at (7.9,4.7) {$e$};
 \node at (9.3, 6) {$\frac{\lc}{\lc+d+e}$};
 \node at (9.3, 5) {$\frac{d  }{\lc+d+e}$};
 \node at (9.3, 4) {$\frac{e  }{\lc+d+e}$};

 \fill (10.2, 5) circle[radius=2pt];
 \draw (10.2, 5) -- (11.2,5.3);
 \node at (11.8, 5.3) {$\frac{\lc}{\lc+e}$};
 \node at (10.7,5.3) {$c$};

 \draw (10.2, 5) -- (11.2,4.7);
 \node at (11.8, 4.7) {$\frac{e}{\lc+e}$};
 \node at (10.7,4.7) {$e$};

 \fill (10.2, 4) circle[radius=2pt];
 \draw (10.2, 4) -- (11.2,3.9);
 \node at (11.8, 3.9) {$\frac{\lc}{\lc+e}$};
 \node at (10.7,4.1) {$c$};

 \draw (10.2, 4) -- (11.2,3.1);
 \node at (11.8, 3.1) {$\frac{e}{\lc+e}$};
 \node at (10.7,3.7) {$e$};

 \draw (4, 0) -- (5,1);
 \node at (6, 1)  {$\frac{a}{a+b+c+e}$};

 \draw (4, 0) -- (5,0);
 \node at (6, 0)  {$\frac{b}{a+b+c+e}$};

 \draw (4, 0) -- (5,-1);
 \node at (6, -1)  {$\frac{c}{a+b+c+e}$};

 \draw (4, 0) -- (5,-2);
 \node at (6, -2)  {$\frac{e}{a+b+c+e}$};

 \node at (11.8, 0) {$\frac{\lb}{\lb+\lc}$};
 \draw (10.2, -1) -- (11.2,-1);

 \node at (11.8, -1) {$\frac{c}{\lb+\lc}$};

 \draw (10.2, -2) -- (11.2,-1.6);
 \node at (11.8, -1.6) {$\frac{\la}{\la+\lc}$};

 \draw (10.2, -2) -- (11.2,-2.2);
 \node at (11.8, -2.2) {$\frac{\lc}{\la+\lc}$};

 \draw (10.2, -3) -- (11.2,-2.9);
 \node at (11.8, -2.9) {$\frac{\la}{\la+\lb}$};

 \draw (10.2, -3) -- (11.2,-3.4);
 \node at (11.8, -3.4) {$\frac{\lb}{\la+\lb}$};

 \node at (9.3, 2) {$\frac{\lb}{\lb+\lc+e}$};
 \node at (9.3, 1) {$\frac{\lc}{\lb+\lc+e}$};
 \node at (9.3, 0) {$\frac{e  }{\lb+\lc+e}$};

 \fill (10.2, 1) circle[radius=2pt];  % da finishes; paths bce
 \draw (10.2, 1) -- (11.2,2);
 \node at (11.8, 2) {$\frac{\lb}{\lb+e}$};
 \draw (10.2, 1) -- (11.2,1);
 \node at (11.8, 1) {$\frac{e}{\lb+e}$};

```{tikz, label=tikz3,fig.cap = "Partial probability tree for five competitors $a$-$e$ with mutually supporting subsets $\\left\\lbrace a,b,c\\right\\rbrace$ [with support term $\\lambda$] and $(de)$ [with support term $\\mu$], hyper3 approach", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw (0,0) -- (2,3); 
\draw (0,0) -- (2,2);
\draw (0,0) -- (2,1);
\draw (0,0) -- (2,0);
\draw (0,0) -- (2,-1);
\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}\underbrace{\left\lbrace d,e\right\rbrace}_{\mu}$};

@@

@@

```{r calcmaxmax}
maxHmax <- maxp(RB_BB_LF(s = osup$maximum))
```

>>>>

---
title: "Function `cheering()` in the `hyper2` package"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
---

@@

To cite the `hyper2` package in publications, please use @hankin2017_rmd.

This document is a distillation and clarification of
`inst/plackett_luce_monster.Rmd`.

In the following, I use the ''race'' metaphor: five runners take part
in a race and arrive in order: $a\succ b\succ c\succ d\succ 5e.
Consider the following Plackett-Luce likelihood function for this
observation:

\begin{equation}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_b}{      p_b + p_c + p_d + p_e}\cdot
\frac{p_c}{            p_c + p_d + p_e}\cdot
\frac{p_d}{                  p_d + p_e}\cdot
\frac{p_e}{                        p_e}
\end{equation}

Note the use of multiplication between the terms, indicating
conditional independence.  We may use the multiplicative
generalization of Bradley-Terry strengths to introduce an element of
non-independence between the terms.  There are many ways to do this,
but one simple case would be to define equivalence classes of the
competitors with each equivalence class comprising mutually supporting
runners.  Equivalence classes of size one correspond to unsupported
runners.  The metaphor would be that a runner who has finished the
race is able to support other members of his equivalence class by
cheering his teammates, boosting their better performance.  The
`hyper2` package includes function `cheering()` which implements this
functionality.

As an example, consider figure 0.1 This shows a partial probability
tree diagram for some of the $5!=120$ possible order statistics.  The
standard Plackett-Luce likelihoods have been modified to account for
two groups of mutually supporting runners: $\left\lbrace
a,b,c\right\rbrace$ with support term $\lambda$, and $\left\lbrace
d,e\right\rbrace$ with support term $\mu$.  From START, the first
runner to cross the finishing line is associated with standard
Plackett-Luce.  Taking the top path as an example, we see that the
likelihood function for $a\succ b\succ c\succ d\succ e$ would be

```{tikz, label=tikz3, fig.cap = "Partial probability\\label{tikzabcde} tree for five competitors $a$-$e$ with mutually supporting subsets $\\left\\lbrace a,b,c\\right\\rbrace$ [with support term $\\lambda$] and $(de)$ [with support term $\\mu$], hyper3 approach", fig.ext = 'png', echo=FALSE}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\fill (0,0) circle[radius=2pt]; % root; paths abcde
\draw (0,0) -- (2,3);
\draw (0,0) -- (2,2);
\draw (0,0) -- (2,1);
\draw (0,0) -- (2,0);
\draw (0,0) -- (2,-1);
\node at (0,2.5) (eq1) {$\underbrace{\left\lbrace a,b,c\right\rbrace}_{\lambda}\underbrace{\left\lbrace d,e\right\rbrace}_{\mu}$};

@@

\begin{equation}\label{asuccb}
\frac{p_a}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{\lambda p_b}{\lambda p_b + \lambda p_c + p_d + p_e}\cdot
\frac{\lambda p_c}{\lambda p_c + p_d + p_e}\cdot
\frac{p_d}{p_d + p_e}
\end{equation}

In equation \ref{asuccb}, the first term is standard Plackett-Luce: at
this point, nooone has finished and cheering effects are absent.  The
second term reflects the fact that competitors $b$ and $c$ are
supported by competitor $a$, who has by this point finished the race
and is supporting his teammates.

By contrast, the likelihood function for observation $d\succ a\succ
c\succ b\succ e$ would be

\begin{equation}
\frac{p_d}{p_a + p_b + p_c + p_d + p_e}\cdot
\frac{p_a}{p_a + p_b + p_c + \mu p_e}\cdot
\frac{\lambda p_c}{\lambda p_b + \lambda p_c + p_e}\cdot
\frac{\lambda p_b}{\lambda p_b + \mu p_e}
\end{equation}

where this likelihood function reflects the mutual support for
equivalence class $\left\lbrace d,e\right\rbrace$.  Note that the
final term reflects the fact that competitors $b$ and $e$ have their
support active when vying for fourth place, as members of both their
teams have finished at this point.

This probability model could easily be modified to account for
specific circumstances.  The cheering effect be asymmetrical (with $a$
helping $b$ but not vice-versa, for example).  The effect could
operate only on specific ordered pairs.  Or perhaps the effect might
have a finite lifetime: if $a$ places $n^\mathrm{th}$, then the
cheering effect is active only for competitors placing
$(n+r)^\mathrm{th}$ or above.

@@

```{r calcmaxmax, cache=TRUE}
maxHmax <- maxp(RB_BB_LF(s = osup$maximum))
```
